Notes on Prediction Types

### 2011-06-27 docs/pert_notes/pert.tex
list of tables
add details about advective form vs. conservative form in make_edge_scal

\section{General form of an advection equation}
In predicting the interface states, we work with an advection-like
equation, and call anything other than the advective term a ``force''.
For example, for a scalar $\phi$, obeying a conservation law with a source term, $f$:
\begin{equation}
\frac{\partial \phi}{\partial t} = -\nabla \cdot (\phi \Ub) + f
\end{equation}
We write this in advective form as:
\begin{equation}
\frac{\partial \phi}{\partial t} = -\Ub \cdot \nabla \phi \, \underbrace{- \phi \nabla \cdot \Ub}_{\text{``advective force''}} + f
\end{equation}
Here we identify the non-advection portion of the divergence of the
flux as the ``advective force'' for the advective equation.  This
advective force will be used as a source when predicting the edge
states.  Depending on the nature of $f$, it may either appear in the
prediction as well or be absent if the process it describes is
incorporated by Strang-splitting.

@@

\begin{table*}[h]
\centering
\caption{Forcing term into {\tt make\_edge\_scal} \newline}
\label{table:pred:hforce}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{l|c}
\hline
\hline
{\tt enthalpy\_pred\_type} &   {advective force} \\
\hline
1 / {\tt predict\_rhohprime} $((\rho h)^\prime)$ &  
 $-(\rho h)^\prime \; \nabla \cdot (\Ubt+w_0 \er) - 
 \nabla \cdot (\Ubt (\rho h)_0) + (\Ubt \cdot \er) \frac{\partial p_0}{\partial r} + \nabla \cdot \kth \nabla T$ \\
2 / {\tt predict\_h}  $(h)$ & $\frac{1}{\rho} \left [\psi + (\Ubt \cdot \er)
  \frac{\partial p_0}{\partial r} + \nabla \cdot \kth \nabla T \right ]$ \\
3 / {\tt predict\_T\_then\_rhohprime} $(T)$ & $\frac{1}{\rho c_p} \left \{ (1 - \rho h_p) 
   \left [\psi + (\Ubt \cdot \er) \frac{\partial p_0}{\partial r} \right ] + \nabla \cdot \kth \nabla T \right \}$ \\
4 / {\tt predict\_T\_then\_h}  $(T)$ & $\frac{1}{\rho c_p} \left\{ (1 - \rho h_p) \left [\psi + (\Ubt \cdot \er)
\frac{\partial p_0}{\partial r}\right ] +  \nabla \cdot \kth \nabla T \right\}$ \\
\hline
\end{tabular}
\end{table*}

@@

\begin{table*}[h]
\centering
\caption{EOS states in {\tt makeHfromRhoT\_edge}\newline}
\label{table:pred:EOSinputs}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{l|c|c}
\hline
\hline
{\tt species\_pred\_type} & {$\rho$ edge state} &   {$X_k$ edge state}  \\
\hline
{\tt predict\_rhoprime\_and\_X} & 
  $\rho_0^{n+\myhalf,\rm{avg}} + \rho'_\mathrm{edge}$ & 
  $(X_k)_\mathrm{edge}$ \\
{\tt predict\_rhoprime\_and\_rhoX} & 
  $\sum_k (\rho X_k)_\mathrm{edge}$ & 
  $(\rho X_k)_\mathrm{edge}/\sum_k (\rho X_k)_\mathrm{edge}$ \\
{\tt predict\_rho\_and\_X} & 
  $\rho_\mathrm{edge}$ & 
  $(X_k)_\mathrm{edge}$ \\
\hline
\end{tabular}
\end{table*}

>>>>

\section{Predicting interface states}

The \maestro\ hyperbolic equations come in two forms: advective and conservative.
The procedure for predicting interface states differs slightly depending on which
form we are dealing with.

\subsection{Advective form}

Most of the time, we are dealing with equations in advective form.
Here, a scalar, $\phi$, obeys:
\begin{equation}
\frac{\partial \phi}{\partial t} = -\Ub \cdot \nabla \phi + f
\end{equation}
where $f$ is the force.  This is the form that the perturbation
equations take, as well as the equation for $X_k$ itself.

A piecewise linear prediction of $\phi$ to the interface 
would appear as:
\begin{eqnarray}
\phi_{i+1/2,j}^{n+1/2} &=& \phi_{i,j}^n 
    + \left . \frac{\Delta x}{2} \frac{\partial \phi}{\partial x} \right |_{i,j}
    + \left . \frac{\Delta t}{2} \frac{\partial \phi}{\partial t} \right |_{i,j} \\
 &=& \phi_{i,j}^n 
    + \left . \frac{\Delta x}{2} \frac{\partial \phi}{\partial x} \right |_{i,j}
    +  \frac{\Delta t}{2} \left ( -u \frac{\partial \phi}{\partial x} 
                                         -v \frac{\partial \phi}{\partial y} + f \right ) \\
 &=& \phi_{i,j}^n + \frac{\Delta x}{2} \left ( 1 - \frac{\Delta t}{\Delta x} u \right ) 
           \frac{\partial \phi}{\partial x} 
    \underbrace{- \frac{\Delta t}{2} v \frac{\partial \phi}{\partial y}}_{\mathrm{``transverse~term''}} + \frac{\Delta t}{2} f
\end{eqnarray}
(see the Godunov notes section for more details).  Here, the
``transverse term'' is accounted for in {\tt make\_edge\_scal}.  Any
additional forces should be added to $f$.  For the perturbation form
of equations, we add additional advection-like terms to $f$ by calling
{\tt modify\_scal\_force}.  This will be noted below.

\subsection{Conservative form}

A conservative equation takes the form:
\begin{equation}
\frac{\partial \phi}{\partial t} = -\nabla \cdot ( \phi \Ub) + f
\end{equation}
%
Now a piecewise linear prediction of $\phi$ to the interface is 
\begin{eqnarray}
\phi_{i+1/2,j}^{n+1/2} &=& \phi_{i,j}^n 
    + \left . \frac{\Delta x}{2} \frac{\partial \phi}{\partial x} \right |_{i,j}
    + \left . \frac{\Delta t}{2} \frac{\partial \phi}{\partial t} \right |_{i,j} \\
 &=& \phi_{i,j}^n 
    + \left . \frac{\Delta x}{2} \frac{\partial \phi}{\partial x} \right |_{i,j}
    +  \frac{\Delta t}{2} \left ( -\frac{\partial (\phi u)}{\partial x} 
                                  -\frac{\partial (\phi v)}{\partial y} + f \right ) \\
 &=& \phi_{i,j}^n + \frac{\Delta x}{2} \left ( 1 - \frac{\Delta t}{\Delta x} u \right ) 
           \frac{\partial \phi}{\partial x} 
    \underbrace{- \frac{\Delta t}{2} \phi \frac{\partial u}{\partial x} 
                - \frac{\Delta t}{2} \frac{\partial (\phi v)}{\partial y}}_{\mathrm{``transverse~term''}} + \frac{\Delta t}{2} f
\end{eqnarray}
Here the ``transverse term'' includes the non-advective portion of the $x$-flux (for the $x$-prediction).
Both of the underbraced terms are accounted for in {\tt make\_edge\_scal} automatically when
we call it with {\tt is\_conservative = .true.}.

@@

\begin{table*}[h]
\centering
\caption{Forcing term into {\tt make\_edge\_scal}}
\label{table:pred:hforce}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{l|c}
\hline
\hline
{\tt enthalpy\_pred\_type} &   {advective force} \\
\hline
1 / {\tt predict\_rhohprime} $((\rho h)^\prime)$ &  
 $-(\rho h)^\prime \; \nabla \cdot (\Ubt+w_0 \er) - 
 \nabla \cdot (\Ubt (\rho h)_0) + (\Ubt \cdot \er) \frac{\partial p_0}{\partial r} + \nabla \cdot \kth \nabla T$ \\
2 / {\tt predict\_h}  $(h)$ & $\frac{1}{\rho} \left [\psi + (\Ubt \cdot \er)
  \frac{\partial p_0}{\partial r} + \nabla \cdot \kth \nabla T \right ]$ \\
3 / {\tt predict\_T\_then\_rhohprime} $(T)$ & $\frac{1}{\rho c_p} \left \{ (1 - \rho h_p) 
   \left [\psi + (\Ubt \cdot \er) \frac{\partial p_0}{\partial r} \right ] + \nabla \cdot \kth \nabla T \right \}$ \\
4 / {\tt predict\_T\_then\_h}  $(T)$ & $\frac{1}{\rho c_p} \left\{ (1 - \rho h_p) \left [\psi + (\Ubt \cdot \er)
\frac{\partial p_0}{\partial r}\right ] +  \nabla \cdot \kth \nabla T \right\}$ \\
\hline
\end{tabular}
\end{table*}

@@

\begin{table*}[h]
\centering
\caption{EOS states in {\tt makeHfromRhoT\_edge}}
\label{table:pred:EOSinputs}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{l|c|c}
\hline
\hline
{\tt species\_pred\_type} & {$\rho$ edge state} &   {$X_k$ edge state}  \\
\hline
{\tt predict\_rhoprime\_and\_X} & 
  $\rho_0^{n+\myhalf,\rm{avg}} + \rho'_\mathrm{edge}$ & 
  $(X_k)_\mathrm{edge}$ \\
{\tt predict\_rhoprime\_and\_rhoX} & 
  $\sum_k (\rho X_k)_\mathrm{edge}$ & 
  $(\rho X_k)_\mathrm{edge}/\sum_k (\rho X_k)_\mathrm{edge}$ \\
{\tt predict\_rho\_and\_X} & 
  $\rho_\mathrm{edge}$ & 
  $(X_k)_\mathrm{edge}$ \\
\hline
\end{tabular}
\end{table*}
